---
title: "Quiz2"
author: "mgestal"
date: "Wednesday, January 14, 2015"
output: html_document
---

To run successfully the Quiz, next packages should be previously installed:

   - install.packages("sqldf")
   - install.packages("httpuv")


```{r}

## Load required packages
library("data.table")
library("sqldf")

## create directory for downloaded data
if(!file.exists("data")) {
  dir.create("data")
}

````

### Question 1

Register an application with the Github API here  https://github.com/settings/applications. 

Access the API to get information on your instructors repositories (hint: this 
is the url you want "https://api.github.com/users/jtleek/repos"). Use this data 
to find the time that the datasharing repo was created. What time was it created? 

This tutorial may be useful (https://github.com/hadley/httr/blob/master/demo/oauth2-github.r). 
You may also need to run the code in the base R package and not R studio.


**IMPORTANT: Execute from RConsole not from RStudio**

library(httr)
require(httpuv)
require(jsonlite)


1. Find OAuth settings for github:
    http://developer.github.com/v3/oauth/

oauth_endpoints("github")

2. Register an application at https://github.com/settings/applications;
    Use any URL you would like for the homepage URL (http://github.com is fine)
    and http://localhost:1410 as the callback url

    Insert your client ID and secret below - if secret is omitted, it will
    look it up in the GITHUB_CONSUMER_SECRET environmental variable.
myapp <- oauth_app("github", key="a200ea9915500822d8ea", secret = "51f2a63ad2d120deefe4a3e9e3dd600defc851fd")

3. Get OAuth credentials
github_token <- oauth2.0_token(oauth_endpoints("github"), myapp)

4. Use API
req <- GET("https://api.github.com/users/jtleek/repos", config(token = github_token))
stop_for_status(req)
output <- content(req)


printRepo <- function(name, owner, created) {
  msg <- sprintf("name: %25s   owner: %6s   created: %10s", name, owner, created)
  message(msg)
}

for(n in 1:length(output)) {
  printRepo(output[[n]]$name, output[[n]]$owner[[1]], output[[n]]$created)
}

 output: 

- name:                  ballgown   owner: jtleek   created: 2013-08-28T18:18:50Z
- name:   capitalIn21stCenturyinR   owner: jtleek   created: 2014-05-27T20:38:31Z
- name:              dataanalysis   owner: jtleek   created: 2013-01-22T12:16:41Z
- name:             datascientist   owner: jtleek   created: 2012-06-24T14:36:20Z
- name:               datasharing   owner: jtleek   created: 2013-11-07T13:25:07Z
- name:        datawomenontwitter   owner: jtleek   created: 2014-09-09T15:49:57Z
- name:                 derfinder   owner: jtleek   created: 2013-07-20T16:05:59Z
- name:               derfinder-1   owner: jtleek   created: 2013-11-08T18:16:27Z
- name:                       DSM   owner: jtleek   created: 2013-08-08T13:20:40Z
- name:               EDA-Project   owner: jtleek   created: 2013-03-12T14:46:24Z
- name:             futureofstats   owner: jtleek   created: 2013-11-21T15:07:59Z
- name:            genomicspapers   owner: jtleek   created: 2014-10-07T18:32:10Z
- name:                googleCite   owner: jtleek   created: 2013-09-03T12:48:28Z
- name:                  graduate   owner: jtleek   created: 2012-11-07T16:43:18Z
- name:                 healthvis   owner: jtleek   created: 2012-11-12T21:35:30Z
- name:                      hyde   owner: jtleek   created: 2014-04-25T03:59:31Z
- name:           inclassfeb62014   owner: jtleek   created: 2014-02-06T16:13:11Z
- name:                  jhsph753   owner: jtleek   created: 2013-01-22T03:06:39Z
- name:              jhsph753and4   owner: jtleek   created: 2014-01-04T21:06:44Z
- name:          jtleek.github.io   owner: jtleek   created: 2014-04-25T04:14:45Z
- name:                      leah   owner: jtleek   created: 2014-03-05T16:11:46Z
- name:                  leekasso   owner: jtleek   created: 2014-01-08T02:40:21Z
- name:                   modules   owner: jtleek   created: 2013-07-12T21:09:41Z
- name:                  MOOCtalk   owner: jtleek   created: 2014-03-13T13:53:05Z
- name:                 polyester   owner: jtleek   created: 2014-06-06T01:57:33Z
- name:                 portfolio   owner: jtleek   created: 2014-01-21T13:35:16Z
- name:                 protocols   owner: jtleek   created: 2014-08-21T20:55:43Z
- name:               rdsmGeneSig   owner: jtleek   created: 2013-08-15T13:16:01Z
- name:                   reviews   owner: jtleek   created: 2013-10-22T23:47:38Z
- name:                   rfitbit   owner: jtleek   created: 2013-12-30T17:43:29Z

Solution: 
  
  printRepo(output[[4]]$name, output[[4]]$owner[[1]], output[[4]]$created)

output

- name:               datasharing   owner: jtleek   created: 2013-11-07T13:25:07Z


## Question 2

 The sqldf package allows for execution of SQL commands on R data frames. 
 We will use the sqldf package to practice the queries we might send with the 
 dbSendQuery command in RMySQL. Download the American Community Survey data and 
 load it into an R object called acs

 [https://d396qusza40orc.cloudfront.net/getdata%2Fdata%2Fss06pid.csv](https://d396qusza40orc.cloudfront.net/getdata%2Fdata%2Fss06pid.csv) 

Which of the following commands will select only the data for the probability weights pwgtp1 with ages less than 50?        

```{r}

if (!file.exists("./data/ss06pid.csv")) {
    fileUrl <- "https://d396qusza40orc.cloudfront.net/getdata%2Fdata%2Fss06pid.csv" 
    destination <- file.path(getwd(), "data", "ss06pid.csv")
    download.file(fileUrl, destination)
} 

acs <- read.csv("./data/ss06pid.csv")

#query <- sqldf("select pwgtp1 from acs")                        ## NO
#query <- sqldf("select * from acs where AGEP < 50 and pwgtp1")  ## NO
#query <- sqldf("select * from acs where AGEP < 50")             ## NO
query <- sqldf("select pwgtp1 from acs where AGEP < 50")

```

##  Question 3

 Using the same data frame you created in the previous problem, what is the 
 equivalent function to unique(acs$AGEP)

```{r}
unique(acs$AGEP)
query1 <- sqldf("select distinct AGEP from acs")
# > query2 <- sqldf("select AGEP where unique from acs")   ## NO
# > query3 <- sqldf("select unique * from acs")            ## NO
# > query4 <- sqldf("select unique AGEP from acs")  

# query4: Error in sqliteSendQuery(con, statement, bind.data) : 
#  error in statement: near "unique": syntax error
identical(query1$AGEP, unique(acs$AGEP))

```

## Question 4

 How many characters are in the 10th, 20th, 30th and 100th lines of HTML 
 from this page: 
  
 [http://biostat.jhsph.edu/~jleek/contact.html](http://biostat.jhsph.edu/~jleek/contact.html) 

```{r}

connection <- url("http://biostat.jhsph.edu/~jleek/contact.html")
htmlCode <- readLines(connection)
close(connection)

# htmlCode contais one html line per element index
nchar(htmlCode[c(10,20,30,100)])

```

##  Question 5


Read this data set into R and report the sum of the numbers in the fourth column.

[https://d396qusza40orc.cloudfront.net/getdata%2Fwksst8110.for](https://d396qusza40orc.cloudfront.net/getdata%2Fwksst8110.for)

Original source of the data: [http://www.cpc.ncep.noaa.gov/data/indices/wksst8110.for](http://www.cpc.ncep.noaa.gov/data/indices/wksst8110.for)
(Hint this is a fixed width file format)

**Note:** downloading and reading from url, gives me an error in Rmd file. That commands are replaces with calls that use the local data file

```{r}

url <- "https://d396qusza40orc.cloudfront.net/getdata%2Fwksst8110.for"

if (!file.exists("./data/Fwksst8110.for")) {
  download.file(url, "./data/Fwksst8110.for")
}

rawData <- read.csv("./data/Fwksst8110.for", header = TRUE)

#lines <- readLines(url, n = 10)
lines <- readLines(file("./data//Fwksst8110.for", open="r") , n = 10)


w <- c(-1,9,-5,4,4,-5,4,4,-5,4,4,-5,4,4)
colNames <- c("week", "sstNino12", "sstaNino12", "sstNino3", "sstaNino3", "sstNino34", 
              "sstaNino34", "sstNino4", "sstaNino4")
#d <- read.fwf(url, w, header = FALSE, skip = 4, col.names = colNames)
d <- read.fwf("./data/Fwksst8110.for", w, header = FALSE, skip = 4, col.names = colNames)

sum(d[, 4])

```