---
title: "Quiz1"
author: "mgestal"
date: "Wednesday, January 14, 2015"
output: html_document
---


To run successfully the Quiz, next packages should be previously installed:

  - install.packages("jpgeg")
  - install.packages("jsonlite")
  - install.packages("data.table")
  - install.packages("httr")
  - install.packages("xlsx")
  - install.packages("xml")


``` {r}

## load packages
library("data.table")
library("xlsx")
library("XML")

## create directory for downloaded data
if(!file.exists("data")) {
  dir.create("data")
}

```

##  Question 1

The American Community Survey distributes downloadable data about United 
States communities. Download the 2006 microdata survey about housing for the 
state of Idaho using download.file() from here:
  
[https://d396qusza40orc.cloudfront.net/getdata%2Fdata%2Fss06hid.csv](https://d396qusza40orc.cloudfront.net/getdata%2Fdata%2Fss06hid.csv)

load the data into R. The code book, describing the variable names is here:
  
[https://d396qusza40orc.cloudfront.net/getdata%2Fdata%2FPUMSDataDict06.pdf](https://d396qusza40orc.cloudfront.net/getdata%2Fdata%2FPUMSDataDict06.pdf)

How many housing units in this survey were worth more than $1,000,000?

```{r}

if (!file.exists("./data/housing.csv")) {
    fileUrl <- "https://d396qusza40orc.cloudfront.net/getdata%2Fdata%2Fss06hid.csv"
    download.file(fileUrl, destfile="./data/housing.csv")
}

if (!file.exists("./data/PUMSDataDict06.pdf")) {
    dictUrl <- "https://d396qusza40orc.cloudfront.net/getdata%2Fdata%2FPUMSDataDict06.pdf"
    f <- file.path(getwd(), "data", "PUMSDataDict06.pdf")
    download.file(dictUrl, destfile=f, mode = "wb")
}


```

VAL Property = 24 if price > $1,000,000

option 1

```{r}

housing <- read.csv("./data/housing.csv")
sum(!is.na(housing$VAL[housing$VAL==24]))

```


option 2 : using DataTables

```{r}

DT <- data.table(read.csv("./data/housing.csv"))

tables()
setkey(DT, VAL)
DT[, .N, by=VAL]  # or DT[, .N, key(DT)]


```

## Question 2

Use the data you loaded from Question 1. Consider the variable FES in the code 
book. Which of the "tidy data" principles does this variable violate?


Solution >  Tidy data has one variable per column.  
Explanation > FES references to Family type and employment status


## Question 3

Download the Excel spreadsheet on Natural Gas Aquisition Program here: 
 
 [https://d396qusza40orc.cloudfront.net/getdata%2Fdata%2FDATA.gov_NGAP.xlsx](https://d396qusza40orc.cloudfront.net/getdata%2Fdata%2FDATA.gov_NGAP.xlsx) 

Read rows 18-23 and columns 7-15 into R and assign the result to a variable called "dat"

What is the value of:   sum(dat\$Zip*dat$Ext, na.rm=T) 


```{r}

library("xlsx")

# It should be executed inside RStudio, because it produces an error on Rmd if file does not exist
if (!file.exists("./data/ngap.xlsx")) {
    fileUrl <- "https://d396qusza40orc.cloudfront.net/getdata%2Fdata%2FDATA.gov_NGAP.xlsx"
    destination <-file.path(getwd(), "data", "ngap.xlsx")
    download.file(fileUrl, destination, mode="wb")
}

rows <- 18:23
columns <- 7:15

dat<-read.xlsx("./data/ngap.xlsx", sheetIndex=1, colIndex=columns, rowIndex=rows)
sum(dat$Zip * dat$Ext, na.rm=T)

```

## Question 4

Read the XML data on Baltimore restaurants from here:
  
[https://d396qusza40orc.cloudfront.net/getdata%2Fdata%2Frestaurants.xml](https://d396qusza40orc.cloudfront.net/getdata%2Fdata%2Frestaurants.xml)

How many restaurants have zipcode 21231?

- option 1

```{r}

fileUrl <- "https://d396qusza40orc.cloudfront.net/getdata%2Fdata%2Frestaurants.xml"

# It should be executed inside RStudio, because it produces an error on Rmd if file does not exist
if (!file.exists("./data/restaurants.xml")) {
    destination <-file.path(getwd(), "data", "restaurants.xml")
    download.file(fileUrl, destination)
}

#restaurants <- xmlTreeParse(fileUrl, useInternal=TRUE)
restaurants <- xmlTreeParse(file="./data/restaurants.xml", useInternal=TRUE)
rootNode <- xmlRoot(restaurants)
rootNode[[1]][[1]]

zipCodes <- xpathSApply(rootNode, "//zipcode", xmlValue)
length(zipCodes[zipCodes==21231])  # or sum(xpathSApply(rootNode, "//zipcode", xmlValue)==21231)

```

- option 2

```{r}

table(zipCodes == 21231)

```

## Question 5

 The American Community Survey distributes downloadable data about United States 
 communities. Download the 2006 microdata survey about housing for the state of 
 Idaho using download.file() from here:
   
   [https://d396qusza40orc.cloudfront.net/getdata%2Fdata%2Fss06pid.csv](https://d396qusza40orc.cloudfront.net/getdata%2Fdata%2Fss06pid.csv)

 using the fread() command load the data into an R object DT

 Which of the following is the fastest way to calculate the average value of the variable pwgtp15
 broken down by sex using the data.table package?


```{r}

if (!file.exists("./data/ss06pid.csv")) {
    fileUrl <- "https://d396qusza40orc.cloudfront.net/getdata%2Fdata%2Fss06pid.csv"
    download.file(fileUrl, destFile="./data/ss06pid.csv")
}

DT <- fread("./data/ss06pid.csv")

check <- function(y, t) {
  message(sprintf("Elapsed time: %.10f", t[3]))
  print(y)
}

t <- system.time(y <- sapply(split(DT$pwgtp15, DT$SEX), mean))
check(y, t)

t <- system.time(y <- mean(DT$pwgtp15, by = DT$SEX))
check(y, t)

t <- system.time(y <- DT[, mean(pwgtp15), by = SEX])
check(y, t)

t <- system.time(y <- mean(DT[DT$SEX == 1, ]$pwgtp15)) + system.time(mean(DT[DT$SEX ==  2, ]$pwgtp15))
check(y, t)

t <- system.time(y <- tapply(DT$pwgtp15, DT$SEX, mean))
check(y, t)

```