---
title: 'Regression Models: Project'
author: "Marcos Gestal"
output:
  html_document:
    keep_md: yes
---

## Context

You work for Motor Trend. Looking at a data set of a collection of cars, they are interested in exploring the relationship between a set of variables and miles per gallon (MPG) (outcome). They are particularly interested in the following two questions:

* "Is an automatic or manual transmission better for MPG"
* "Quantify the MPG difference between automatic and manual transmissions"

```{r echo=FALSE}
data(mtcars)
mtcars$am <- factor(mtcars$am, labels=c("Automatic", "Manual"))
```


## Data summary

```{r}
automaticCars <- mtcars[mtcars$am=="Automatic", ]
manualCars <- mtcars[mtcars$am=="Manual", ]

summary(mtcars$mpg)
summary(automaticCars$mpg)
summary(manualCars$mpg)
```

See Appendix 1 for a graphical view of the relation between MPG and type of transmission (boxplot).

These informations (boxplot and mean values in summaries) seem to show a best performance about mpg values for Automatic transmissions.


## Manual vs Automatic Transmission

### Statistical inference: T Test

A t-Test is performed to check wether there is a statistically significant difference between the mgp values for automatic and manual transmissions. The p-value < 0.05 so the null hypothesis can be reject, so we would accept that **there are differences in the mpg values due the type of transmissions**. The t-test results give us information about **MPG mean for automatic cars is lower than the MPG for manual cars**. It confirms our previous assumption about the best performance of automatic transmissions. Now a study will be done to quantify this difference.


```{r}
t.test(manualCars$mpg, automaticCars$mpg)
```

### Linear Regression

```{r}
summary( lm(mpg ~ am, data = mtcars) )
```

A brief study of the data shows how for automatic transmissions, the slope 7.245 for manual transmissions vs. automatic ones, so the performace is better (lower) for the last ones. The problem of this study relies in that it only explains 36% of the variance (Multiple R-Squared value)

We repeat the linear regression model, but now with the intercepted term in the origin

```{r}
summary( lm(mpg ~ 0 + am, data = mtcars) )
```

Here we can observe how the performance for automatic transmissions is better than the manual ones.

### Multivariate Regression

In this case, we extend the regression model to include other directly relevant variables to explain the
mpg value

```{r}
 advancedModel <- lm(mpg ~ am + gear + wt + hp + cyl + carb, data = mtcars)
 summary ( lm(mpg ~ am + gear + wt + hp + cyl + carb, data = mtcars) )
```

This new regression model explains the 85.67% of the variance. We can see how the manual transmission and the number
of gears increases the mpg values (transmission in a higher percentage). 


```{r}
simpleModel <- lm(mpg ~ 0 + am, data = mtcars)
advancedModel <- lm(mpg ~ am + gear + wt + hp + cyl + carb, data = mtcars)

anova(simpleModel, advancedModel)

```

## Conclussions

We performed a statistical study of the data and we can conclude that MPG for automatic cars presents lower values than the MPG for the manual cars.

## Appendices

### Appendix 1: boxplot

```{r echo=FALSE, fig.height=3.5}
boxplot(mpg ~ am, data = mtcars, col = "salmon", 
        main = "MPG by Transmission Type",
        xlab="Transmission", ylab = "MPG (Miles Per Gallon)")
```

### Appendix 2

```{r echo=FALSE}

# par(mfrow=c(2, 2))
# plot(simpleModel)

par(mfrow=c(2, 2))
plot(advancedModel)
```

